OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA

### 2019-01-31 phil_paper/paper_wjv.tex
new motivation (post-stef)

\usepackage[notablist,nofiglist]{endfloat}
%\renewcommand{\efloatseparator}{\mbox{}}

@@

	Households often share connections to public utilities with their neighbors in order to split set-up and maintenance costs.  This paper analyzes how sharing connections affects optimal pricing for water utilities.  With microdata from 1.5 million water connections in Manila, I estimate water demand in the presence of water sharing and solve for prices that maximize consumer surplus while funding production costs.  In practice, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  Instead, the optimal policy uses high fixed prices to cover costs and low marginal prices to encourage greater consumption.  Sharing households split fixed prices and enjoy inexpensive piped water with their neighbors instead of using low-quality, non-piped water.  Counterfactual exercises find increases in shared connections, total piped water consumption, and social welfare from optimal prices compared to current prices in Manila.

@@

Faced with high fixed costs of purchasing and maintaining durable goods, consumers often share these goods with each other.  While online platforms have accelerated peer-to-peer rentals of cars, apartments, and storage space (\cite{horton2016owning}), households in developing countries often share connections to public utilities like water and electricity with their neighbors.\footnote{70\% of African water utilities report frequent sharing of water taps between households.  Over 14\% of urban households in Arba-Minch, Ethiopia report sourcing electricity from their neighbors instead of purchasing directly from the utility (\cite{mustefa2016energy}).  In South Africa as of 2012, 11.9\% of the population report sharing their mobile phone with others.  This percentage jumps to 19.9\% among those living in poverty (\cite{calandro2012mobile}).}  Sharing can have large consequences for optimal pricing, particularly for public utilities.  Public utility pricing is especially salient in developing countries because prices must often be set to fully cover costs (\cite{laffont2005regulation}) and because prices determine not only monthly consumption, but also which households can access these basic services.  For example, when households cannot share connections, they may prefer relatively low fixed prices (ie. installation and maintenance fees) so that each household can afford a connection and high marginal prices (ie. prices per gallon or watt) to cover production costs.  However, when households can easily share connections, they may prefer splitting high fixed prices with their neighbors and enjoying low marginal prices from shared connections.  
%The optimal pricing policy depends both on consumer demand and on the costs of sharing connections between neighbors.

This paper focuses empirically on pricing piped water in Manila, Philippines to increase access and welfare when households share connections.  Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  Yet, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, any subsidies to public utilities are often limited by high costs of public funds and risk of regulatory capture in developing countries (\cite{laffont2005regulation}).

@@

Insecure land-tenure, credit-constraints, and administrative costs often make it difficult for disadvantaged households to purchase their own water connections (\cite{devoto2012happiness}).  As a result, households frequently share connections ranging from fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) to drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of these  sharing networks, research has not yet examined the importance of water sharing for optimal pricing policy.

@@

These estimates allow me to solve a standard regulator pricing problem: setting prices to cover fixed and marginal costs while maximizing consumer surplus.  I also let the regulator place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which have relatively lower marginal prices, zero upfront costs, and varying water quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.  Without fixed prices, land-tenure requirements, or administrative expenses, these sources often have low costs of access.}  The opposite approach --- high fixed fees and low marginal prices --- maximizes both access to piped water and welfare, even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements enable small-users to access piped water from their neighbors for relatively lower marginal prices than those offered by non-piped sources.  By reducing the total number of connections, sharing arrangements also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.\footnote{In this regulatory arrangement, cost savings are passed on to consumers through lower prices.}

@@

Finally, I compute a tariff that maximizes consumer surplus only for the bottom 50\% of users and is intended to capture goals of ensuring minimum usage and access as well as income redistribution.  With sharing, this tariff is ineffective at reallocating surplus from large to small users; however in a counterfactual without sharing, this tariff achieves substantial redistribution.  These findings suggest that informal sharing networks may constrain the extent to which non-linear prices can redistribute surplus.

@@

This paper focuses on one of the providers which serves 1.5 million connections and around 5 million people.  Figure~\ref{figure:tariff} shows the regulated tariff in Manila.  The increasing marginal price is thought to cross-subsidize consumption among low-demand, poorer households through higher bills from large users.  The regulator assigns zero marginal price to the first 10 cubic meters of monthly usage, which is called the ``lifeline'' rate.  This policy is intended to provide a minimum level of consumption for all households.  This tariff may also perform efficient price discrimination if low-usage households are more price-sensitive.
\begin{figure}
\centering
\caption{Marginal Prices in Manila}\label{figure:tariff}
\begin{center}
\includegraphics[scale=.7]{tables/tariff_structure.pdf}\\
\footnotesize{1 USD = 50 PhP}
\end{center}
\end{figure}
% detail more about how to get water, with the fixed fee and conne{}ction fee?!
On top of this tariff schedule, the regulator also charges a fixed fee independent of usage totaling 225 PhP/month.  This fee includes both the monthly service charge and the initial connection fee that I have put into monthly terms for the duration of the water connection.\footnote{Appendix~\ref{appendix:connectionfeeamortization} provides more details about how this monthly fee is calculated and amortized.}  The fixed fee allows the provider to recover any operating costs associated with serving the connection.  This form of increasing tariff and fixed fee combination is implemented in over 70\% of developing cities as well as over 50\% of developed cities (\cite{hoque2013state}).

@@

The empirical setting in Manila provides a broad range of price variation to identify these price sensitivity terms separately from the fixed preferences.  First, the non-linear pricing structure given by Figure~\ref{figure:tariff} forces households to decide whether to consume above or below tariff kink points as they face monthly shocks to their demand.  Second, regulated tariff changes over time affecting all households generate time-series variation in the tariff, which can be seen in Appendix~\ref{appendix:tariffvariation}.  Third, I use rate-changes which affect individual connections differently over time described in more detail in Appendix~\ref{appendix:ratechangeelasticity}.  The structural model combines all of these sources of price variation in the estimation.  The identification assumption is that the consumption and preference shocks, $\epsilon,\eta$ are normally distributed and uncorrelated with these price changes.

@@

\section{Connection Fee}\label{appendix:connectionfeeamortization}

This section constructs a total fee paid to the water provider each month as the sum of the monthly service fee, 150 PhP/month on average, and the connection fee of 6,000 PhP, amortized over the duration of the water connection.  To amortize the connection fee without information on full durations of residence for all households, I assume that households move to Manila, use a water connection for 80 months, then leave the city.  This assumption is roughly consistent with the hazard rate of disconnection for water connections.  Dividing the connection fee by 80 months and adding it to the monthly service fee results in a total monthly fee of 225 PhP.  Any credit constraints, discount rates, or other barriers to accessing a water connection are captured in monthly terms in the fixed cost estimates in Section~\ref{section:fixedcostestimation}.

>>>>

%\usepackage[notablist,nofiglist]{endfloat}

%\renewcommand{\efloatseparator}{\mbox{}}

@@

%This paper analyzes how sharing connections affects optimal pricing for water utilities.

	Households often share connections to public utilities with their neighbors in order to split set-up and maintenance costs.  With microdata from 1.5 million water connections in Manila, this paper estimates water demand when connections are shared and solves the regulator's problem: setting prices that maximize consumer surplus while covering production costs.  In practice, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  Instead, the optimal policy covers production costs with high fixed prices, which are split between sharing households, and sets low marginal prices to encourage consumption.  Counterfactual exercises find increases in shared connections, total piped water consumption, and social welfare from the optimal policy compared to current prices in Manila.

%Sharing households split fixed prices with their neighbors to access the low marginal prices.

@@

Faced with high fixed costs of purchasing and maintaining durable goods, consumers often share these goods with each other.  While online platforms have accelerated peer-to-peer rentals of cars, apartments, and storage space (\cite{horton2016owning}), households in developing countries often share connections to public utilities like water and electricity with their neighbors.\footnote{70\% of African water utilities report frequent sharing of water taps between households.  Over 14\% of urban households in Arba-Minch, Ethiopia report sourcing electricity from their neighbors instead of purchasing directly from the utility (\cite{mustefa2016energy}).  In South Africa as of 2012, 11.9\% of the population report sharing their mobile phone with others.  This percentage jumps to 19.9\% among those living in poverty (\cite{calandro2012mobile}).}  Sharing can have large consequences for optimal pricing, particularly for public utilities.  This paper focuses empirically on pricing for piped water in Manila, Philippines in order to increase access and welfare when households share connections.

Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  Yet, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, any subsidies to public utilities are often limited by high costs of public funds and risk of regulatory capture in developing countries (\cite{laffont2005regulation}).  In many cases, prices must be set to fully recover costs.

%Instead of piped water, these households often resort to local deepwells or water vendors.  CITE THE ADB HERE!

%Public utility pricing is especially salient in developing countries because prices must often be set to fully cover costs (\cite{laffont2005regulation}) and because prices determine not only monthly consumption, but also which households can access these basic services.  For example, when households cannot share connections, they may prefer relatively low fixed prices (ie. installation and maintenance fees) so that each household can afford a connection and high marginal prices (ie. prices per gallon or watt) to cover production costs.  However, when households can easily share connections, they may prefer splitting high fixed prices with their neighbors and enjoying low marginal prices from shared connections.  
%The optimal pricing policy depends both on consumer demand and on the costs of sharing connections between neighbors.

@@

Insecure land-tenure, credit-constraints, and administrative costs often make it difficult for disadvantaged households to purchase their own water connections (\cite{devoto2012happiness}).  As a result, households frequently share connections including fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) and drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of these  sharing networks, research has not yet examined the importance of water sharing for optimal pricing policy.

@@

These estimates allow me to solve a standard regulator pricing problem: setting prices to cover fixed and marginal costs while maximizing consumer surplus.  I also let the regulator place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which have relatively lower marginal prices, zero upfront costs, and varying water quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.}  The opposite approach --- high fixed fees and low marginal prices --- maximizes both access to piped water and welfare, even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements enable small-users to access piped water from their neighbors for relatively lower marginal prices than those offered by non-piped sources.  By reducing the total number of connections, sharing arrangements also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.\footnote{In this regulatory arrangement, cost savings are passed on to consumers through lower prices.}  
%  Without fixed prices, land-tenure requirements, or administrative expenses, these sources often have low costs of access.

@@

Finally, I compute a tariff that maximizes consumer surplus only for the bottom 50\% of users and is intended to capture goals of ensuring minimum usage and access as well as income redistribution.  With sharing, this tariff is ineffective at reallocating surplus from large to small users; however in a counterfactual without sharing, this tariff achieves substantial redistribution.  These findings suggest that informal sharing networks may constrain the extent to which non-linear prices can redistribute consumer surplus.

@@

This paper focuses on one of the providers which serves 1.5 million connections and around 5 million people.  Figure~\ref{figure:tariff} shows the regulated tariff in Manila.\footnote{Marginal prices continue increasing gradually at 20 m3 intervals above 50 m3, which are not pictured in Figure~\ref{figure:tariff}.  Appendix~\ref{appendix:connectionfeeamortization} includes the full tariff as presented to consumers.}  The increasing marginal price is thought to cross-subsidize consumption among low-demand, poorer households through higher bills from large users.  The regulator assigns zero marginal price to the first 10 cubic meters of monthly usage, which is called the ``lifeline'' rate.  This policy is intended to provide a minimum level of consumption for all households.  This tariff may also perform efficient price discrimination if low-usage households are more price-sensitive.
\begin{figure}
\centering
\caption{Tariff in Manila}\label{figure:tariff}
\begin{center}
\includegraphics[scale=.7]{tables/tariff_structure.pdf}\\
\footnotesize{Figure includes average marginal prices from 2010 to 2015. \\  1 USD $\sim$ 50 PhP}
\end{center}
\end{figure}
% detail more about how to get water, with the fixed fee and conne{}ction fee?!
On top of this tariff schedule, the regulator also charges a fixed fee independent of usage totaling 225 PhP/month.  This fee includes both the monthly service charge and the initial connection fee that I have put into monthly terms for the duration of the water connection.\footnote{Appendix~\ref{appendix:connectionfeeamortization} provides more details about how this monthly fee is calculated and amortized.}  The fixed fee allows the provider to recover any operating costs associated with serving the connection.  This form of increasing tariff and fixed fee combination is implemented in over 70\% of developing cities as well as over 50\% of developed cities (\cite{hoque2013state}).

@@

The empirical setting in Manila provides a broad range of price variation to identify these price sensitivity terms separately from the fixed preferences.  First, the non-linear pricing structure given by Figure~\ref{figure:tariff} forces households to decide whether to consume above or below tariff kink points as they face monthly shocks to their demand.  Although the full tariff schedule includes seven segments as described in Figure~\ref{figure:tarifftrue} in Appendix~\ref{appendix:connectionfeeamortization}, the model includes only the first four tariff segments, applying the fourth tariff segment to all greater consumption levels.  Any biases caused by this assumption may be limited both because households rarely consume above 60 m3 per month and because marginal prices increase gradually at higher levels of consumption.  Also, the estimation abstracts away from a discrete increase in the fixed fee that occurs for consumption above 10 m3 per month as shown in Figure~\ref{figure:tarifftrue} in Appendix~\ref{appendix:connectionfeeamortization}.  Taken together, these limitations may lead the model to overestimate price elasticities since some households may be responding to higher marginal prices than those included in the model.

Second, regulated tariff changes over time affecting all households generate time-series variation in the tariff, which can be seen in Appendix~\ref{appendix:tariffvariation}.  Third, I use rate-changes which affect individual connections differently over time described in more detail in Appendix~\ref{appendix:ratechangeelasticity}.  The structural model combines all of these sources of price variation in the estimation.  The identification assumption is that the consumption and preference shocks, $\epsilon,\eta$ are normally distributed and uncorrelated with these price changes.

@@

\section{Tariff and Fee Structure}\label{appendix:connectionfeeamortization}

\begin{figure}
\caption{Residential Tariff Presented to Consumers}\label{figure:tarifftrue}
\begin{center}
\includegraphics[scale=.8]{tables/tariff_pic_png.png} \\
\footnotesize{``conn.'' refers to connection, ``cu.m.'' refers to cubic meters of usage \\ per month, and \textbf{P} refers to Philippine Pesos.}
\end{center}
\end{figure}

Figure~\ref{figure:tarifftrue} includes the current tariff as it is presented to consumers.  The average total fee paid to the water provider each month includes the sum of the monthly service fee, 150 PhP/month on average between 2010 and 2015 (including water meter replacement charges and disconnection and reconnection fees not listed in Figure~\ref{figure:tarifftrue}), and the connection fee of 6,000 PhP, amortized over the duration of the water connection.  
%Although the fixed fee listed on the bill increases for households consuming above 10 m3 as indicated in Figure~\ref{figure:tarifftrue}, the empirical strategy assumes one fixed fee at this higher level
%includes the fixed fee for households using above 10 m3 in the calculation since the bulk of

To amortize the connection fee without information on full durations of residence for all households, I assume that households move to Manila, use a water connection for 80 months, then leave the city.  This assumption is roughly consistent with the hazard rate of disconnection for water connections.  Dividing the connection fee by 80 months and adding it to the monthly service fee results in a total monthly fee of 225 PhP.  Any credit constraints, discount rates, or other barriers to accessing a water connection are captured in monthly terms in the fixed cost estimates in Section~\ref{section:fixedcostestimation}.
